State of the Practice for Lattice Boltzmann Method Software

### 2022-01-18 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revision of text connecting RQs

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice} such as
reflective and non-reflective boundaries. They consider the behavior of a
collection of particles as a single unit at the mesoscopic scale, between the
nanoscopic and microscopic scales. LBM solvers predict the positional
probability of a collection of particles moving through a lattice structure
following a two step process: i) streaming, where the particles
move along the lattice via links; and, ii) colliding, where energy and
momentum is transferred among particles that collide \citep{bao2011lattice}. As
an example of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds numbers.
\wss{Zahra might have a better example?  A journal publication isn't likely
going to accept a figure produced for another paper.}  LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Given the value of their time, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for LBM libraries, the time
is right to assess the state of the practice.  The goal of this report is to
analyze the current state of LBM software development to provide insight into
its best practices and to offer guidance for future development.

To focus our efforts, we devised five research questions, which can be applied
to LBM, as well as to other domains, of research software \citep{SmithEtAl2021}.
The first question below refers to artifacts, which is the term we use to
represent the documents, scripts and code that constitutes a software
development project. Example artifacts include requirements, specifications,
user manuals, unit tests, system tests, usability tests, build scripts, API
(Application Programming Interface) documentation, READMEs, license documents,
process documents, and code.

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What LBM
	software projects exist, with the constraint that the source code must be
	available for all identified projects?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to
	the artifacts present in their repositories?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with
	respect to the use of tools for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do LBM projects compare to research software in general with respect to
	principles, processes and methodologies used?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects? 
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers:
	\begin{enumerate} 
		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
		address these concerns?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}).
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in \rqref{RQ_PainPoints} and the concerns identified in \rqref{RQ_Concerns}.a?

@@

\subsubsection{Lack of Development Time}

@@

\subsubsection{Lack of Software Development Experience}

@@

\subsubsection{Lack of Incentive and Funding}

@@

\subsubsection{Lack of External Support}

@@

\subsubsection{Parallelization and Continuous Integration}

@@

\subsubsection{Ensuring Correctness}

@@

\wss{** BEGIN Testing and Verification **}
\citet{Nguyen-HoanEtAl2010} says the following:
\begin{itemize}
	\item 38 of 60 respondents performed unit tests
	\item 23 of 60 respondents performed integration tests
	\item ``In our results, the four most common activities involve comparisons
	with known, trusted, or prior data. Verification and peer review were three
	of the four least commonly performed activities.'' \wss{testing is not the
	only way to verify software.  This addresses the oracle problem.}
\end{itemize}
\wss{** END Testing and Verification **}

@@

\subsubsection{Technical Debt}

@@

\subsubsection{Quality of Documentation}

@@

Answers \rqref{RQ_Recommend}.  This section answers the research questions
listed in Section~\ref{secIntro} using the quantitative data presented in
Section~\ref{AHPresults}, qualitative data presented in
Section~\ref{interviewresults}, additional data from software package
repositories and artifacts, and domain expert feedback. Each subsection answers
one of the research questions. Software package artifacts are examined in
Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
principles, processes, and methodologies are discussed in Section
~\ref{prinprocmeth}, development pain points are explored in Section
~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
(Section~\ref{bestPractices}) and future practices
(Section~\ref{futurePractices}). Finally, threats to the validity of this
assessment are noted in Section~\ref{threats}.

@@

	\item Installability. Include OS compatibility, including specific OS
	versions. This was done by several top ranked packages, including ESPResSo
	and LUMA.
	\item Installability. Provide complete installation instructions. This was
	done by all of the top five ranked packages.
	\item Installability. Installation instructions should be written as if the
	user does not have any dependencies installed and is installing on a clean
	OS. This was done by several top ranked packages, including ESPResSo,
	OpenLB, and Palabos. This can be tested on a clean environment using a
	virtual machine.
	\item Installability. List all dependencies in the installation
	instructions, dependency versions, and how to install them. This was done by
	several top ranked packages, including ESPResSo, LUMA, OpenLB, and Palabos.
	\item Installability. If possible, automate the installation of
	dependencies. Use tools such as Make. This was done by all of the top five
	ranked packages.
	\item Installability. Automate the installation process as much as possible.
	Use tools such as Make. This was done by all of the top five ranked
	\item Installability. The installation instruction should only be in one
	location. This was done by several top ranked packages, including ESPResSo,
	OpenLB, and Palabos.
	\item Installability. Include descriptive error messages for errors
	encountered during installation. This was done by several top ranked
	packages, including Ludwig and LUMA.
	\item Installability. Provide a way to validate the installation. This can
	be done using a custom script, or a test case that specifies expected
	output. This was done by several top ranked packages, including OpenLB,
	Ludwig, LUMA, and Palabos.
	\item Installability. Provide instructions for uninstalling the software.
	These were included with pyLBM.

>>>>

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice}, such as
reflective and non-reflective boundaries. LBM considers the behavior of a
collection of particles as a single unit at the mesoscopic scale, which lies
between the nanoscopic and microscopic scales. LBM solvers predict the
positional probability of a collection of particles moving through a lattice
structure following a two step process: i) streaming, where the particles move
along the lattice via links; and, ii) colliding, where energy and momentum is
transferred among particles that collide \citep{bao2011lattice}. As an example
of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds
numbers. \wss{We should ask Zahra and her students for a better example.}  LBM
has several advantages over conventional CFD methods, including a simple
calculation procedure, improved parallelization, and robust handling of complex
geometries \citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Given the value of their time, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for LBM libraries, the time
is right to assess the state of the practice.  The goal of this report is to
analyze the current state of LBM software development and compare it to the
development practices employed for research software in general.  We want to
highlight the success stories that LBM developers can share amongst their
community and with the broader research software community.  We also want to
identify areas for future improvement.

To focus our efforts, we devised 11 research questions.  The questions are
inspired by the questions that underpin our data collection methodology
\citep{SmithEtAl2021}. To start with, we need to identify the relevant examples
of LBM software.  Knowing that our data collection methodology \citep{SmithEtAl2021} requires access to source code, we build that constraint into our research question:

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What LBM
	software projects exist, with the constraint that the source code must be
	available for all identified projects?
\end{enumerate}

We next wish to assess the representative software to determine how well they
apply current software development best practices.  At this point we will base
our assessment on publicly available artifacts, where artifacts are the
documents, scripts and code that we find in a project's repository. Example
artifacts include requirements, specifications, user manuals, unit test cases,
system tests, usability tests, build scripts, API (Application Programming
Interface) documentation, READMEs, license documents, process documents, and
code.  Following best practices does not guarantee popularity, so we will also
estimate which projects are most often used by the scientists.

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
\end{enumerate}

To understand the state of the practice we wish to learn the frequency with
which different artifacts appear, the types of development tools used and the
methodologies used for software development.  With this data, we can ask
questions about how the LBM community compares to other research software.  That
is, in what ways does LBM follow the trends from other communities, and in which
ways is it different?

\begin{enumerate}
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do LBM projects compare to research software in general with respect to
	the artifacts present in their repositories?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do LBM projects compare to research software in general with
	respect to the use of tools for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do LBM projects compare to research software in general with respect to
	principles, processes and methodologies used?
\end{enumerate}

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to talk to developers.  We need to learn their
concerns, how they deal with these concerns, and what pain points exist for
them.  We wish to know what practices are used by the top LBM projects, so that
others can potentially emulate these practices. We also wish to identify new
practices by borrowing successful ideas from other domains. For LBM developers
that may be interested in cutting edge software practices, ideas that are
proposed, but not yet in practice, should also be identified.  All of this is
covered by the questions outlined below:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects? 
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general?
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers:
	\begin{enumerate} 
		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
		address these concerns?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}).
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in \rqref{RQ_PainPoints} and the concerns identified in \rqref{RQ_Concerns}.a?

@@

\subsection{Lack of Development Time}

@@

\subsection{Lack of Software Development Experience}

@@

\subsection{Lack of Incentive and Funding}

@@

\subsection{Lack of External Support}

@@

\subsection{Parallelization and Continuous Integration}

@@

\subsection{Ensuring Correctness}

@@

\wss{** BEGIN Testing and Verification **}

\citet{Nguyen-HoanEtAl2010} says the following:
\begin{itemize}
	\item 38 of 60 respondents performed unit tests
	\item 23 of 60 respondents performed integration tests
	\item ``In our results, the four most common activities involve comparisons
	with known, trusted, or prior data. Verification and peer review were three
	of the four least commonly performed activities.'' \wss{testing is not the
	only way to verify software.  This addresses the oracle problem.}
\end{itemize}

May do more testing than artifacts indicate, but that is still a problem.
Artifacts should be available so that others can verify the results, verify
their installation.  Part of an assurance case means that a third party can
verify the work.

\wss{** END Testing and Verification **}

@@

\subsection{Technical Debt}

@@

\subsection{Quality of Documentation}

@@

Answers \rqref{RQ_Recommend}.  This section answers the research questions
listed in Section~\ref{secIntro} using the quantitative data presented in
Section~\ref{AHPresults}, qualitative data presented in
Section~\ref{interviewresults}, additional data from software package
repositories and artifacts, and domain expert feedback.

@@

	\item The top projects provide great examples of achieving installability,
	by taking such steps as: providing complete installation instructions,
	writing instructions as if the user does not have any of the dependencies
	installed, automation of the installation process (for instance using make),
	including descriptive error message, and including steps for uninstallation.
